\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-08 paper.tex

\subsection{Recombination breakpoint intervals}
\label{sec:breakpoint_intervals}
It is rarely possible to be precise about the position on the genome at which
a recombinant sequence switches from inheriting from one parent to another.
Even if we observe the recombinant sequence before subsequent
divergence occurs (during onward transmission), there is no way to identify the exact breakpoint if the two parent sequences
are similar.
Here we define the
interval \sally{within which a particular breakpoint may have occurred} as the genome coordinates over which the
sequences for the left and right parent nodes are identical. The right-hand
extreme of the breakpoint interval is chosen by the LS HMM Viterbi algorithm
(Section~\ref{sec:ls}), and the left endpoint is then derived by directly
comparing the parent sequences. See Section~\ref{sec:treatment_recombinants} for
a precise definition of breakpoints and their intervals.

@@

It is important to note that the precise endpoints of intervals can be
somewhat arbitrary, because they are defined by \sally{the sequence differences that happen to be present in the} recombinant's parents.
Thus, breakpoint intervals will tend to be truncated at sites that are
hypermutable, either due to increased information about parentage
or spurious inferences caused by sequencing errors.
It is therefore helpful to compare the number of intersecting intervals
with the number of mutations per site (Figure~\ref{fig:breakpoint-distribution}).
For instance, position 28,271 (located in the ORF8-N intergenic
region) has the largest number of mutations and appears as an endpoint
of 66 intervals.
The 7,572 mutations (including 5,782 insertions and 1,605 deletions)
at this site are an indicator that this
homopolymeric region may be prone to sequencing errors
and \sally{potentially} should be included in the list of ``problematic sites'' that
are excluded from analysis (Section~\ref{sec:data_preprocessing}).
On the other hand, 58 of the breakpoint intervals have endpoints
within one base of site 27,972, which has undergone 770 mutation events
(including 425 C$>$T, 314 T$>$C). The C$>$T mutation has the effect of
truncating ORF8, and it has been posited that the truncation is neutral
or advantageous for transmission, and disadvantageous within-host
\citep{Jungreis2021-dh}, suggesting that the high rate of recurrent mutation
at this site may be due to selection.

>>>>

\subsection{Recombination breakpoint intervals}
\label{sec:breakpoint_intervals}
It is rarely possible to be precise about the position on the genome at which
a recombinant sequence switches from inheriting from one parent to another.
Even if we observe the recombinant sequence before subsequent
divergence occurs (during onward transmission), there is no way to identify the exact breakpoint if the two parent sequences
are similar.
Here we define the
interval within which a particular breakpoint may have occurred
as the genome coordinates over which the
sequences for the left and right parent nodes are identical. The right-hand
extreme of the breakpoint interval is chosen by the LS HMM Viterbi algorithm
(Section~\ref{sec:ls}), and the left endpoint is then derived by directly
comparing the parent sequences. See Section~\ref{sec:treatment_recombinants} for
a precise definition of breakpoints and their intervals.

@@

It is important to note that the precise endpoints of intervals can be
somewhat arbitrary, because they are defined by the sequence differences
that happen to be present in the recombinant's parents.
Thus, breakpoint intervals will tend to be truncated at sites that are
hypermutable, either due to increased information about parentage
or spurious inferences caused by sequencing errors.
It is therefore helpful to compare the number of intersecting intervals
with the number of mutations per site (Figure~\ref{fig:breakpoint-distribution}).
For instance, position 28,271 (located in the ORF8-N intergenic
region) has the largest number of mutations and appears as an endpoint
of 66 intervals.
The 7,572 mutations (including 5,782 insertions and 1,605 deletions)
at this site are an indicator that this
homopolymeric region may be prone to sequencing errors
and potentially should be included in the list of ``problematic sites'' that
are excluded from analysis (Section~\ref{sec:data_preprocessing}).
On the other hand, 58 of the breakpoint intervals have endpoints
within one base of site 27,972, which has undergone 770 mutation events
(including 425 C$>$T, 314 T$>$C). The C$>$T mutation has the effect of
truncating ORF8, and it has been posited that the truncation is neutral
or advantageous for transmission, and disadvantageous within-host
\citep{Jungreis2021-dh}, suggesting that the high rate of recurrent mutation
at this site may be due to selection.
